---
title: "Microscopy Metadata"
teaching: 45
exercises: 20
---

:::::::::::::::::::::::::::::::::::::: questions

- What is metadata?
- Why is metadata important in microscopy?
- How do pixel measurements relate to physical measurements?
- How can metadata be accessed in R?
- What is the difference between an Image and an AnnotatedImage?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Define metadata and explain its importance.
- Identify common microscopy metadata fields.
- Explain the relationship between pixels and physical units.
- Read an OME-TIFF image using RBioFormats.
- Compare the Image and AnnotatedImage classes.
- Recognise why metadata preservation is essential for reproducible research.

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

In the previous lesson we learned that microscopy images are stored in image
files such as TIFF and OME-TIFF.

The image pixels themselves are only part of the story.

To interpret image data correctly we must also understand the information that
describes how the image was acquired.

This information is known as **metadata**.

Metadata are often described as:

`Data about data`

In microscopy, metadata can be just as important as the pixel data.

## What Is Metadata?

Consider a digital photograph taken with a smartphone.

The image consists of pixels, but the file may also contain information such as:

- camera model
- date and time
- GPS location
- exposure settings

These are examples of metadata. Photos use a metadata model known as EXIF.

Microscopy images are similar.

The image contains pixel values, while metadata describe how those pixel values
were acquired.

::::::::::::::::::::::::::::::::::::: challenge

Which of the following would typically be considered metadata?

1. A pixel intensity value
2. A microscope objective lens
3. A detector exposure time
4. A channel name

:::::::::::::::::::::::: solution

The correct answers are:

```text
2. A microscope objective lens
3. A detector exposure time
4. A channel name
```

Pixel intensities are image data rather than metadata.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

## Common Microscopy Metadata

Modern microscopes record large amounts of metadata during image acquisition.

Examples include:

### Image Information

- image dimensions
- bit depth
- number of channels
- number of z-planes
- number of time points

### Spatial Information

- pixel size
- voxel dimensions

### Instrument Information

- microscope type
- objective magnification
- numerical aperture
- detector type

### Acquisition Settings

- exposure time
- detector gain
- laser wavelength
- laser power

### Experimental Information

- sample identifier
- treatment condition
- acquisition date
- operator

::::::::::::::::::::::::::::::::::::: callout

## Metadata Enable Reproducibility

Without metadata it is often impossible to determine precisely how an image was
acquired.

Good metadata are essential for reproducible and trustworthy image analysis.

::::::::::::::::::::::::::::::::::::::::::::::::

## Pixels Are Not Physical Units

Suppose a nucleus occupies:

```text
5000 pixels
```

What is its area?

The answer is:

```text
We do not know.
```

Pixels are not physical units.

To convert pixels into measurements such as square micrometres we must know the
pixel size.

## Pixel Size

Suppose an image has:

```text
Pixel size = 0.5 µm
```

Each pixel represents:

```text
0.5 µm × 0.5 µm
```

The area represented by a single pixel is:

```{r}
pixel_size <- 0.5

pixel_area <- pixel_size^2

pixel_area
```

## Converting Pixels to Physical Area

Suppose a nucleus occupies:

```text
4000 pixels
```

and the pixel size is:

```text
0.5 µm
```

We can estimate its physical area.

```{r}
pixel_size <- 0.5

n_pixels <- 4000

area_um2 <- n_pixels * pixel_size^2

area_um2
```

::::::::::::::::::::::::::::::::::::: challenge

An object occupies:

```text
2500 pixels
```

The image pixel size is:

```text
0.2 µm
```

Calculate the object's area in µm².

:::::::::::::::::::::::: solution

Area of one pixel:

```text
0.2 × 0.2 = 0.04 µm²
```

Total area:

```text
2500 × 0.04 = 100 µm²
```

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

## Why Metadata Matter

Consider two images.

### Image A

```text
Pixel size = 1.0 µm
```

### Image B

```text
Pixel size = 0.1 µm
```

In both images a nucleus occupies:

```text
5000 pixels
```

The nuclei do not have the same physical area.

Without metadata we cannot determine the real size of either object.

## OME-TIFF: Images Plus Metadata

The Open Microscopy Environment (OME) developed OME-TIFF to provide a
standardised way of storing microscopy images together with their metadata.

Conceptually an OME-TIFF file contains:

```text
Image Pixels
+
Metadata
```

Metadata stored within an OME-TIFF file may include:

- pixel size
- voxel dimensions
- objective information
- channel names
- detector settings
- exposure times
- acquisition dates

Unlike JPEG, OME-TIFF was specifically designed for scientific imaging.

::::::::::::::::::::::::::::::::::::: callout

## Why OME-TIFF Matters

OME-TIFF preserves both the image and the information required to interpret the
image correctly.

::::::::::::::::::::::::::::::::::::::::::::::::

## Introducing Bio-Formats

Modern microscopes use many proprietary file formats.

Examples include:

- Zeiss `.czi`
- Nikon `.nd2`
- Leica `.lif`
- Olympus `.vsi`

Reading these formats directly can be challenging because each stores image data
and metadata differently.

The Open Microscopy Environment developed **Bio-Formats** to solve this problem.

Bio-Formats provides a common interface for reading image data and metadata from
many microscopy formats.

The R package **RBioFormats** provides access to Bio-Formats from within R.

::::::::::::::::::::::::::::::::::::: callout

## Why Bio-Formats Matters

Bio-Formats allows scientists to work with images from many microscope vendors
without requiring vendor-specific software.

::::::::::::::::::::::::::::::::::::::::::::::::

## Reading an OME-TIFF with RBioFormats

Load RBioFormats:

```{r}
library(RBioFormats)
```

Read an OME-TIFF file.

```{r}
bf_img <- read.image(
  "../data/04-microscopy-metadata/example.ome.tif"
)
```

Determine the class of the returned object.

```{r}
class(bf_img)
```

Unlike EBImage, RBioFormats typically returns an **AnnotatedImage** object.

## Exploring Metadata

Because metadata remain associated with the image, we can inspect them directly.

```{r}
metadata(bf_img)
```

The exact information available depends on the original image file.

Examples may include:

- image dimensions
- pixel sizes
- channel names
- acquisition timestamps
- objective information
- detector settings

::::::::::::::::::::::::::::::::::::: challenge

Use:

```{r}
metadata(bf_img)
```

How many different kinds of metadata can you identify in your image?

:::::::::::::::::::::::: solution

The exact answer depends on the image used.

Common examples include image dimensions, physical pixel sizes, acquisition
settings, channel names, and microscope information.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

## AnnotatedImage versus Image

Earlier in this course we used EBImage.

```{r}
library(EBImage)

eb_img <- readImage(
  "../data/04-microscopy-metadata/example.ome.tif"
)
```

Determine its class.

```{r}
class(eb_img)
```

EBImage returns an object of class:

```text
Image
```

The Image object primarily represents image pixel data.

In contrast:

```{r}
class(bf_img)
```

returns:

```text
AnnotatedImage
```

Conceptually:

```text
Image

└── Pixel Array
```

while:

```text
AnnotatedImage

├── Pixel Array
└── Metadata
```

This distinction becomes increasingly important as image-analysis workflows grow
in complexity.

## What Happens When We Read an OME-TIFF with EBImage?

Suppose we use EBImage to load the same OME-TIFF file.

```{r}
img <- readImage(
  "../data/04-microscopy-metadata/example.ome.tif"
)
```

We can inspect:

```{r}
dim(img)
```

```{r}
range(img)
```

```{r}
display(img)
```

The image data are available for analysis.

However, the rich microscopy metadata contained in the original OME-TIFF file
are not generally exposed through the EBImage Image object itself.

## Image Data and Metadata Become Separated

It is useful to think of this process as:

```text
OME-TIFF File

├── Image Pixels
└── Metadata

    ├── Pixel Size
    ├── Objective Lens
    ├── Exposure Time
    ├── Channel Names
    └── Instrument Settings
```

↓

```text
EBImage Image Object

└── Pixel Array
```

The image data have been loaded successfully, but the metadata are no longer
directly associated with the image object.

::::::::::::::::::::::::::::::::::::: challenge

You receive an OME-TIFF image from a collaborator and wish to access:

- image pixels
- pixel size
- channel names
- acquisition metadata

Which package is likely to be most useful?

1. EBImage
2. RBioFormats

:::::::::::::::::::::::: solution

```text
2. RBioFormats
```

RBioFormats is designed to access both image data and associated microscopy
metadata through Bio-Formats.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

## Losing Metadata

Metadata can be lost when images are:

- exported to inappropriate formats
- copied into presentation software
- converted between file formats
- separated from the original acquisition files

Consider the workflow:

```text
Microscope Image
        |
        V
     OME-TIFF
        |
        V
      JPEG
```

The JPEG image may still be viewable.

However, important metadata may no longer be available.

The image remains.

Critical scientific context may be lost.

::::::::::::::::::::::::::::::::::::: challenge

Why might exporting a microscopy image to JPEG cause problems for future
analysis?

:::::::::::::::::::::::: solution

JPEG files typically retain far less microscopy metadata than formats such as
OME-TIFF.

Important information including calibration, acquisition settings, and channel
information may be lost.

This makes reproducibility and quantitative analysis more difficult.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

## Looking Ahead

Metadata provide the information required to interpret biological images
correctly.

Throughout the remainder of this course we will repeatedly combine:

```text
Image Data
+
Metadata
```

to obtain meaningful biological measurements.

In future lessons we will revisit OME-TIFF, Bio-Formats, and OMERO and explore
how image data and metadata can be managed together throughout the image
analysis lifecycle.

## Key Points

::::::::::::::::::::::::::::::::::::: keypoints

- Metadata describe how an image was acquired.
- Pixel measurements cannot be interpreted without calibration information.
- Pixel size enables conversion from pixels to physical units.
- OME-TIFF stores both image data and microscopy metadata.
- Bio-Formats provides a common interface for reading microscopy image formats.
- RBioFormats can access both image data and metadata.
- EBImage primarily works with image pixel data.
- AnnotatedImage objects retain metadata alongside image data.
- Metadata are essential for reproducible microscopy.
- Exporting images to inappropriate formats can result in metadata loss.

::::::::::::::::::::::::::::::::::::::::::::::::
