---
title: "Batch Processing and Reproducible Workflows"
teaching: 45
exercises: 30
---

:::::::::::::::::::::::::::::::::::::: questions

- How can image analysis workflows be applied to many images?
- Why is reproducibility important?
- How can image measurements be combined into a single dataset?
- How can automated image analysis support biological discovery?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Apply a workflow to multiple images.
- Process groups of experimental and control images.
- Organise measurements into a tidy dataset.
- Visualise differences between experimental conditions.
- Explain the importance of reproducible image-analysis workflows.

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

In the previous lesson we developed a workflow to:

```text
Segment Nuclei
      ↓
Measure Shape
      ↓
Measure Intensity
```

and applied it to a small image stack.

However, microscopy experiments often contain hundreds or thousands of images.

Analysing images manually is:

- slow
- difficult to reproduce
- prone to errors

One of the major advantages of using R is that the same workflow can be applied
consistently to many images.

In this lesson we will analyse a small experiment containing:

```text
4 Control Images
4 Drug-Treated Images
```

and investigate whether the treatment affects nuclear fluorescence intensity.

::::::::::::::::::::::::::::::::::::: callout

## Reproducibility

A reproducible workflow produces the same results every time it is run on the
same data.

Reproducibility is a cornerstone of good scientific practice.

::::::::::::::::::::::::::::::::::::::::::::::::

## The Dataset

Our dataset contains:

```text
data/11-batch-processing/

control_01.tif
control_02.tif
control_03.tif
control_04.tif

drug_01.tif
drug_02.tif
drug_03.tif
drug_04.tif
```

Each image contains fluorescently labelled nuclei.

Display one of the images.

```{r}
library(EBImage)

img <- readImage(
  "data/11-batch-processing/control_01.tif"
)

display(img)
```

Our scientific question is:

> Does the drug treatment increase nuclear fluorescence intensity?

## Listing Image Files

We can obtain a list of image files.

```{r}
files <- list.files(
  "data/11-batch-processing",
  pattern = "\\.tif$",
  full.names = TRUE
)

files
```

This allows our workflow to process every image automatically.

::::::::::::::::::::::::::::::::::::: challenge

Why is it better to generate a file list automatically rather than manually
typing every filename?

:::::::::::::::::::::::: solution

Automatic file discovery:

- saves time
- reduces typing errors
- adapts to changing datasets
- improves reproducibility

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

## Building a Reproducible Workflow

A reproducible workflow applies exactly the same processing steps to every
image.

Our workflow will:

```text
Read Image
      ↓
Threshold
      ↓
Refine Trhshold by Opening
      ↓
Watershed Segmentation
      ↓
Measure Mean Nuclear Intensity
```

## Analysing a Single Image

Let's begin with one image.

```{r}
img <- readImage(
  "data/11-batch-processing/control_01.tif"
)
```

Create a segmentation mask.

```{r}
mask <- img > 0.2
```

Clean the mask.

```{r}
mask <- opening(
  mask,
  makeBrush(
    size = 5,
    shape = "disc"
  )
)
```

Separate touching nuclei.

```{r}
labels <- watershed(
  distmap(mask)
)
```

Measure object intensities.

```{r}
features <- as.data.frame(
  computeFeatures.basic(
    labels,
    img
  )
)
```

Inspect the results.

```{r}
head(features)
```

Each row represents a nucleus.

## Creating a Measurement Function

The same workflow must be applied to every image.

Rather than repeatedly copying and pasting code, we can place the workflow into
a function.

```{r}
measure_nuclei <- function(image_file) {

  img <- readImage(image_file)

  mask <- img > 0.2

  mask <- opening(
    mask,
    makeBrush(
      size = 5,
      shape = "disc"
    )
  )

  labels <- watershed(
    distmap(mask)
  )

  features <- as.data.frame(
    computeFeatures.basic(
      labels,
      img
    )
  )

  features
}
```

Now we can analyse any image using a single command.

```{r}
measure_nuclei(
  files[1]
)
```

::::::::::::::::::::::::::::::::::::: callout

## Why Use Functions?

Functions improve reproducibility because they allow the same workflow to be
applied consistently every time.

::::::::::::::::::::::::::::::::::::::::::::::::

## Processing Multiple Images

Create an empty data frame to store measurements.

```{r}
all_measurements <- data.frame()
```

Loop through every image.

```{r}
for (file in files) {

  measurements <- measure_nuclei(file)

  measurements$image <- basename(file)

  all_measurements <- rbind(
    all_measurements,
    measurements
  )

}
```

Inspect the result.

```{r}
head(all_measurements)
```

We now have measurements from every nucleus in every image.

## Creating Experimental Groups

Create a treatment column.

```{r}
all_measurements$treatment <-
  ifelse(
    grepl("control", all_measurements$image),
    "Control",
    "Drug"
  )
```

Inspect the data.

```{r}
head(all_measurements)
```

Our measurement table now follows tidy-data principles:

- one row = one observation = one nucleus
- one column = one variable

::::::::::::::::::::::::::::::::::::: callout

## Tidy Data

Many R packages expect data to follow a tidy structure.

Image-analysis outputs often become much easier to work with after they are
converted into tidy data.

::::::::::::::::::::::::::::::::::::::::::::::::

## Exploring Mean Nuclear Intensity

Load ggplot2.

```{r}
library(ggplot2)
```

Visualise the distribution of mean nuclear intensities.

```{r}
ggplot(
  all_measurements,
  aes(
    x = b.mean
  )
) +
  geom_histogram(
    bins = 20
  ) +
  labs(
    x = "Mean Nuclear Intensity",
    y = "Count",
    title = "Distribution of Nuclear Intensities"
  )
```

Each bar summarises measurements from many nuclei.

::::::::::::::::::::::::::::::::::::: challenge

What does each observation in this histogram represent?

:::::::::::::::::::::::: solution

Each observation corresponds to one segmented nucleus and its measured mean
intensity.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

## Comparing Experimental Conditions

Visualise mean nuclear intensities by treatment.

```{r}
ggplot(
  all_measurements,
  aes(
    x = treatment,
    y = b.mean
  )
) +
  geom_boxplot()
```

This plot allows us to compare the distributions of measurements between the
experimental groups.

## Summarising the Experiment

Calculate the average intensity for each group.

```{r}
aggregate(
  b.mean ~ treatment,
  data = all_measurements,
  FUN = mean
)
```

This provides a simple summary of the experiment.

::::::::::::::::::::::::::::::::::::: challenge

Which group appears to have higher nuclear intensity?

:::::::::::::::::::::::: solution

The drug-treated group should show higher average nuclear intensity.

This suggests that the treatment increased fluorescence signal.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

## From Measurements to Biological Insight

Throughout this course we have transformed images into measurements.

Conceptually:

```text
Microscopy Images
        ↓
Segmentation
        ↓
Objects
        ↓
Measurements
        ↓
Data Table
        ↓
Biological Interpretation
```

Here we have used the workflow to investigate a biological hypothesis:

```text
Drug Treatment
        ↓
Changes Nuclear Intensity?
```

The answer is no longer based on visual inspection alone.

Instead it is supported by quantitative measurements.

::::::::::::::::::::::::::::::::::::: callout

## Quantitative Biology

Image analysis allows biological questions to be addressed using measurements
rather than subjective observation.

::::::::::::::::::::::::::::::::::::::::::::::::

## Exporting Results

Save the measurements.

```{r eval = FALSE}
write.csv(
  all_measurements,
  "nuclear_measurements.csv",
  row.names = FALSE
)
```

The resulting file can be used for:

- statistical analysis
- plotting
- reporting
- sharing with collaborators

## Looking Back

Throughout this course we have developed a complete image-analysis workflow.

```text
Read Images
      ↓
Metadata
      ↓
Visualisation
      ↓
Colour Channels
      ↓
Enhancement
      ↓
Segmentation
      ↓
Morphological Operations
      ↓
Measurement
      ↓
Batch Processing
      ↓
Biological Insight
```

This workflow forms the foundation of many microscopy image-analysis projects.

## Key Points

::::::::::::::::::::::::::::::::::::: keypoints

- Reproducible workflows apply the same processing steps consistently.
- Functions help automate image-analysis workflows.
- Batch processing allows many images to be analysed efficiently.
- Measurement tables should be organised using tidy-data principles.
- Quantitative measurements can be compared across experimental conditions.
- Image analysis transforms images into biological data.
- Reproducibility is an essential component of scientific image analysis.

::::::::::::::::::::::::::::::::::::::::::::::::
